\section{Approximated replacements}

In the preliminaries, a definition for the exact evaluation of bounds from a bound set $\BoundSet(V)$ was given.
With such an exact evaluation, it is possible to evaluate a bound $b \in \BoundSet(\VSet)$ with respect to a state $\valuation \in \Valuation$ to a value $\exacteval{b}{\valuation} \in \ValueSet$.
In general, we would like to be able to evaluate bounds with respect to arbitrary states $\valuation \in \Valuation$ between a lower bound $\lstate \leq \valuation$ and an upper bound $\ustate \geq \valuation$.

Therefore, we introduce approximated evaluations.
In contrast to an exact evaluation, which replaces variables with a single state $\valuation \in \Valuation$, an approximated evaluation considers two input states $\lstate, \ustate \in \Valuation$.
Then, the bound is evaluated in a way that the result is greater than any exact evaluation with a state $\valuation$ with $\lstate \leq \valuation \leq \ustate$.

We define a generalized principle, which is also applicable to substitutions.

\begin{definition}[Approximated replacement]
  Let $K$ be an arbitrary set of elements with an addition $+: K \times K \rightarrow K$, a negation $-: K \rightarrow K$, a multiplication $\cdot: K \times K \rightarrow K$ and a max operator $\max: K \times K \rightarrow K$.
  For a bound $b \in \BoundSet(\VSet)$ and two functions $\lsub, \usub: \VSet \rightarrow K$, we denote with $\lsubst{b}{\lsub}{\usub}$ and $\usubst{b}{\lsub}{\usub}$ functions of the type $(\VSet \rightarrow K) \times (\VSet \rightarrow K) \rightarrow K$.
  These functions replace the variables of the bound $b$ with values from the functions $\lsub, \usub$ such that the following equations hold.
  \[ \lsubst{\infty}{\lsub}{\usub} = \usubst{\infty}{\lsub}{\usub} = \infty \text{ for } \infty \in \BoundSet(\VSet) \]
  \[ \lsubst{k}{\lsub}{\usub} = \usubst{k}{\lsub}{\usub} = k \text{ for all } k \in \mathbb{N} \subseteq \BoundSet(\VSet) \] 
  \[ \lsubst{v}{\lsub}{\usub} = \lsub(v) \text{ for all } v \in \VSet \subseteq \BoundSet(\VSet) \] 
  \[ \usubst{v}{\lsub}{\usub} = \usub(v) \text{ for all } v \in \VSet \subseteq \BoundSet(\VSet) \] 
  \[ \lsubst{-b}{\lsub}{\usub} = -\usubst{b}{\lsub}{\usub} \text{ for all } b \in \BoundSet(\VSet) \] 
  \[ \usubst{-b}{\lsub}{\usub} = -\lsubst{b}{\lsub}{\usub} \text{ for all } b \in \BoundSet(\VSet) \] 
  \[ \lsubst{b_1 + b_2}{\lsub}{\usub} = \lsubst{b_1}{\lsub}{\usub} + \lsubst{b_2}{\lsub}{\usub} \text{ for all } b_1, b_2 \in \BoundSet(\VSet) \] 
  \[ \usubst{b_1 + b_2}{\lsub}{\usub} = \usubst{b_1}{\lsub}{\usub} + \usubst{b_2}{\lsub}{\usub} \text{ for all } b_1, b_2 \in \BoundSet(\VSet) \] 
  \[ \lsubst{b_1 \cdot b_2}{\lsub}{\usub} = \minimum{
    \usubst{b_1}{\lsub}{\usub} \cdot \usubst{b_2}{\lsub}{\usub},
    \usubst{b_1}{\lsub}{\usub} \cdot \lsubst{b_2}{\lsub}{\usub},
    \lsubst{b_1}{\lsub}{\usub} \cdot \usubst{b_2}{\lsub}{\usub},
    \lsubst{b_1}{\lsub}{\usub} \cdot \lsubst{b_2}{\lsub}{\usub}
  } \text{ for all } b_1, b_2 \in \BoundSet(\VSet) \] 
  \[ \usubst{b_1 \cdot b_2}{\lsub}{\usub} = \maximum{
    \usubst{b_1}{\lsub}{\usub} \cdot \usubst{b_2}{\lsub}{\usub},
    \usubst{b_1}{\lsub}{\usub} \cdot \lsubst{b_2}{\lsub}{\usub},
    \lsubst{b_1}{\lsub}{\usub} \cdot \usubst{b_2}{\lsub}{\usub},
    \lsubst{b_1}{\lsub}{\usub} \cdot \lsubst{b_2}{\lsub}{\usub}
  } \text{ for all } b_1, b_2 \in \BoundSet(\VSet) \] 
  \[ \lsubst{\max(b_1, b_2)}{\lsub}{\usub} = \maximum{\lsubst{b_1}{\lsub}{\usub}, \lsubst{b_2}{\lsub}{\usub}} \text{ for all } b_1, b_2 \in \BoundSet(\VSet) \]
  \[ \usubst{\max(b_1, b_2)}{\lsub}{\usub} = \maximum{\usubst{b_1}{\lsub}{\usub}, \usubst{b_2}{\lsub}{\usub}} \text{ for all } b_1, b_2 \in \BoundSet(\VSet) \]
  \[ \lsubst{k^b}{\lsub}{\usub} = k^{\lsubst{b}{\lsub}{\usub}} \text{ for all } k \in \mathbb{N} \subseteq \BoundSet(\VSet), b \in \BoundSet(\VSet) \]  
  \[ \usubst{k^b}{\lsub}{\usub} = k^{\usubst{b}{\lsub}{\usub}} \text{ for all } k \in \mathbb{N} \subseteq \BoundSet(\VSet), b \in \BoundSet(\VSet) \]
  We call an expression $\lsubst{b}{\lsub}{\usub}$ an \textbf{underapproximated replacement} and $\usubst{b}{\lsub}{\usub}$ an \textbf{overapproximated replacement}.
\end{definition}

The bound set $\BoundSet(\VSet)$ meets the criteria of $K$.
Therefore, two functions $\lsub, \usub: \VSet \rightarrow \BoundSet(\VSet)$ qualify for an approximated replacement.
We call an approximated replacement with such functions an approximated substitution.

\begin{theorem}[Soundness of approximated substitutions]
  Let $\lsub, \usub: \VSet \rightarrow \BoundSet(\VSet)$ be two substitutions with $\exacteval{\lsub(v)}{\valuation} \leq \exacteval{\usub(v)}{\valuation}$ for each variable $v \in \VSet$ and each state $\valuation \in \Valuation$.
  Let $b \in \BoundSet(\VSet)$ be a bound and let $\valuation \in \Valuation$ be a state.
  Then, it holds that
  \[ \exacteval{\lsubst{b}{\lsub}{\usub}}{\valuation} \leq \exacteval{b}{\valuation} \leq \exacteval{\usubst{b}{\lsub}{\usub}}{\valuation}. \]
\end{theorem}

Also, two states $\lstate, \ustate: \VSet \rightarrow \mathbb{Z}$ qualify for an approximated replacement, since $\mathbb{Z}$ meets the criteria of $K$.
We call an approximated replacement with two states an approximated evaluation.

\input{theorems/approximated_evaluations}

The definition of approximated replacements uses implicitly the property of component-wise monotonicity of a bound to infer better over- and underapproximations.

Consider an overapproximated evaluation $\ueval{x}{\lstate}{\ustate}$.
It is easy to see, that the bound $x$ is monotonically increasing in $x$.
Therefore, it is sufficient to substitute the variable $x$ with its overapproximation $\ustate$ to gain a valid overapproximation of the value $\exacteval{x}{\valuation}$ for every state $\valuation$ with $\lstate \leq \valuation \leq \ustate$.

Consider an overapproximated evaluation $\ueval{-x}{\lstate}{\ustate}$.
It is easy to see, that the bound $-x$ is monotonically decreasing in $x$.
Therefore, it is sufficient to substitute the variable $x$ with its underapproximation $\lstate$ to gain a valid overapproximation of the value $\exacteval{-x}{\valuation}$ for every state $\valuation$ with $\lstate \leq \valuation \leq \ustate$.

Consider an overapproximated evaluation $\ueval{x \cdot y}{\lstate}{\ustate}$.
The bound $x \cdot y$ is nor monotonically increasing or decreasing in $x$ neither monotonically increasing or decreasing in $y$.
Therefore, it is necessary to compute all combinations $\usubst{x}{\lstate}{\ustate} \cdot \usubst{y}{\lstate}{\ustate}$, $\usubst{x}{\lstate}{\ustate} \cdot \lsubst{y}{\lstate}{\ustate}$, $\lsubst{x}{\lstate}{\ustate} \cdot \usubst{y}{\lstate}{\ustate}$ and $\lsubst{x}{\lstate}{\ustate} \cdot \lsubst{y}{\lstate}{\ustate}$.
Then, the result can be determined as the maximum of all these combinations.

The following definition formalizes the monotonicity of variables in bounds.

\begin{definition}[Monotonicity of bounds in variables]
  Let $\VSet$ be a finite set of variables.
  Let $b$ be a bound of the bound set $\BoundSet(\VSet)$.
  Let $v \in \VSet$ be a variable.
  We say that two states $\valuation, \valuation' \in \Valuation$ are only different in $v$ if and only if for each variable $v' \in \VSet \setminus \braced{v}$ it holds that $\valuation(v') = \valuation'(v')$ and for the variable $v$ it holds that $\valuation(v) \neq \valuation'(v)$.
  If it furthermore holds that $\valuation(v) > \valuation'(v)$, we call the state $\valuation$ only different and greater in $v$ than the state $\valuation'$.
  If it instead holds that $\valuation(v) < \valuation'(v)$, we call the state $\valuation$ only different and less in $v$ than the state $\valuation'$.
  We call the bound $b$ monotonically increasing in the variable $v$ if and only if for all states $\valuation$ that are only different and greater in $v$ than a state $\valuation'$ it holds that $\exacteval{b}{\valuation} \geq \exacteval{b}{\valuation'}$.
  We call the bound $b$ monotonically decreasing in the variable $v$ if and only if for all states $\valuation$ that are only different and less in $v$ than a state $\valuation'$ it holds that $\exacteval{b}{\valuation} \leq \exacteval{b}{\valuation'}$.
\end{definition}

An approximated replacement yields the better results the more variables a bound is monotonically increasing or decreasing in.
Note that an affine polynomial $b \in \BoundSet_a(\VSet)$ is monotonically increasing or decreasing in all variables.
Therefore, the approximated replacement in an affine polynomial yields better over- and underapproximations than the approximated replacement in non-affine polynomials in $\BoundSet_p(\VSet) \setminus \BoundSet_a(\VSet)$ or arbitrary bounds in $\BoundSet(\VSet)$.
